Using vector space models based on LDA and LSA for automatically grading exam questions

### 2018-06-18 tex/article.tex
Small addition to the discussion.
Rewrote the part of the discussion section where I compared the outcomes of LDA and LSA models. Added a paragraph about comparing models of different dimensionalities.

\begin{abstract}
Students' answers to open-ended exam questions can be automatically graded by comparing them to a `perfect' reference answer in terms of semantic similarity. In this study, I compared three methods for constructing vector spaces in which semantic similarity can be measured. The dimensions of these vector spaces were made up of: 1) Topics, generated by LDA models, versus 2) Singular values, computed by LSA models, versus 3) Vocabulary counts, serving as baseline models. Better results were expected for the LDA and LSA models, as they capture students' answers at the semantic level rather than depending on exact vocabulary matches. Indeed, the best result was found for one of the LDA models, which achieved a Spearman's correlation coefficient of .52 between the grades predicted by the model and the true grades assigned by the lecturer. The LSA models on average outperformed the baseline models as well. In addition, it was found that the type of data used for training the LDA and LSA models impacted the accuracy of the predictions.
\end{abstract}

@@

\section{Discussion}
In this study I compared three types of vector space models for automatically predicting exam grades. The vector space models that used topics (LDA) or singular values (LSA) as their dimensions both outperformed the baseline vector space models that used vocabulary counts as their dimensions. The likely explanation for this is that LDA and LSA models compare documents at the semantic level, rather than looking for exact vocabulary matches \cite{kakkonen2008}. Notably, even in the baseline models the correlation between predicted and lecturer-assigned grades was statistically significant.

Looking back on the literature, \citeasnoun{kakkonen2008} asked whether their LSA models perhaps outperformed their LDA models because they were trained on a small set of student answers only. My results support this explanation: when training on the student answers LSA outperformed LDA, but it was the other way around when using a more diverse document collection (i.e., sentences in a psychology textbook chapter). This raises the question as to why the two types of training data affected the topic models differently (while the size of the training sets was similar).

One explanation can be sought in the optimal number of topics for the LDA models that resulted from a grid search (see Section \ref{sec:training}). This number was different for the two types of training data. In the student answers, the grid search had shown that the optimal number of topics was 2. This is the lowest possible number of topics that an LDA model can have. This outcome makes sense because in essence the student answers all revolved around the same topic (i.e., Whorf's language theory). As a result, however, the resulting vector space models only had two dimensions, which does not allow very fine distinctions between answers of different qualities. When visually inspecting the scatter plots of the outcomes of these models (not printed here), it can be seen that these models predict a 10 for almost all students, resulting in very low correlations. Thus, training on a more diverse corpus such as a textbook seems more suitable when working with LDA models. Indeed, Table 1 shows better results for LDA models that were trained on the textbook and used 20 topics, as compared to LDA models that were trained on student answers.

The textbook-trained LDA models also outperformed the textbook-trained LSA models by a large margin. This is the outcome that Kakkonen et al. (2008) had predicted in their own study. They list several arguments as to why LDA models are expected to function better, including the difficulty to select the right number of dimensions for LSA, the fact that no probability distribution is defined in LSA, and that the reduced matrix in LSA can contain negative values. These explanations likely also apply to the current study. Incidentally, it is nice to see the LDA model achieving the best results, because this is the model whose outcomes are easiest to interpret by students and lecturers.

Zooming in on the individual models, Figures 2-4 show that none of the models ever predicted the grade of 10, even though this was the most common grade in the training set (see Figure 1). Thus, while the lecturer considered many of the students’ answers to be excellent, these answers apparently were quite different from the reference answer in terms of the vector space models that I used for calculating semantic similarity. This shows that \cite{wolfe1998}'s reference answer approach may not be ideal for grading student answers by means of vector space models.

>>>>

\begin{abstract}
Students' answers to open-ended exam questions can be automatically graded by comparing them to a `perfect' reference answer in terms of semantic similarity. In this study, I compared three methods for constructing vector spaces in which semantic similarity can be measured. The dimensions of these vector spaces were made up of: 1) Topics, generated by LDA models, versus 2) Singular values, computed by LSA models, versus 3) Vocabulary counts, serving as baseline models. Better results were expected for the LDA and LSA models, as they capture students' answers at the semantic level rather than depending on exact vocabulary matches. Indeed, the best model was one of the LDA models, which achieved a Spearman's correlation coefficient of .52 between the grades predicted by the model and the true grades assigned by the lecturer. The LSA models on average outperformed the baseline models as well. In addition, it was found that the type of data used for training the LDA and LSA models impacted the accuracy of the predictions.
\end{abstract}

@@

\section{Discussion}
In this study I compared three types of vector space models for automatically predicting exam grades. The vector space models that used topics (LDA) or singular values (LSA) as their dimensions both outperformed the baseline vector space models that used vocabulary counts as their dimensions. The likely explanation for this is that LDA and LSA models compare documents at the semantic level, rather than looking for exact vocabulary matches \cite{kakkonen2008}. Notably, even in all of the baseline models the correlation between predicted and lecturer-assigned grades was statistically significant at the .05 level.

It should be noted that in the results section as well as in the remainder of this discussion, I have made comparisons between models that differed in their dimensionality. For example, the LDA models contained 2 topics when being trained on the student answers, but 20 when being trained on the textbook chapter. These values were chosen because a grid search had shown that they resulted in the most accurate models. However, technically speaking, the outcomes of models that differ in their dimensionality cannot directly be compared, because vectors with lower dimensionalities by default are expected to be more similar. Still, this does not mean that models with lower dimensionalities will \textit{automatically} make the best predictions, which can also be seen from the finding that the best-scoring model was an LDA model with 20 topics. Because I wanted to find the best possible vector space for doing automatic grading, I opted to work with models of different dimensionalities.

\subsection{LDA versus LSA, as a function of training data type}
Looking back on the literature, \citeasnoun{kakkonen2008} asked whether their LSA models perhaps outperformed their LDA models because they were trained on a small set of student answers only. My results support this explanation: when training on the student answers, LSA outperformed LDA, but it was the other way around when using a more diverse document collection (i.e., sentences in a psychology textbook chapter). This raises the question as to why the two types of training data affected the topic models differently (while the size of the training sets was similar).

One explanation can be sought in the optimal number of topics for the LDA models that resulted from the grid search (see Section \ref{sec:training}): 2 when training on student answers, and 20 when training on the textbook chapter. Two is the lowest possible number of topics that an LDA model can have. It is not so unexpected that the grid search yielded this as the best value when training on student answers, as in essence the student answers all revolved around the same topic (i.e., Whorf's language theory). As a result, however, the resulting vector space models only had two dimensions, which does not allow very fine distinctions between answers of different qualities. When visually inspecting the scatter plots of the outcomes of the LDA models with 2 topics trained on student answers, it can be seen that these models predict a 10 for many students, which results in relatively low correlations. As an example, the outcomes of one such model are printed as Figure \ref{lda-predicts-tens} below. For this particular model, $\textit{r}_s$ = .28.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\linewidth]{"LDA predicts tens"}
	\caption{Grades predicted by an LDA model that was trained on the student answers, used raw counting and used the x10 mapping algorithm.}
	\label{lda-predicts-tens}
\end{figure}

Thus, training on a more diverse corpus such as a textbook seems more suitable when working with LDA models. Indeed, Table 1 showed better results for LDA models that were trained on the textbook, as compared to LDA models that were trained on student answers. The textbook-trained LDA models also outperformed the textbook-trained LSA models by a large margin. This is the outcome that \citeasnoun{kakkonen2008} had predicted in their own study. They list several arguments as to why LDA models are expected to function better, including the difficulty to select the right number of dimensions for LSA, the fact that no probability distribution is defined in LSA, and that the reduced matrix in LSA can contain negative values. These explanations likely also apply to the current study. Incidentally, it is nice to see the LDA model achieving the best results, because this is the model whose outcomes are easiest to interpret by students and lecturers.

\subsection{Are our models good enough to be used?}
Zooming in on the individual models, Figures 2-4 show that none of the models ever predicted the grade of 10, even though this was the most common grade in the training set (see Figure 1). Thus, while the lecturer considered many of the students’ answers to be excellent, these answers apparently were quite different from the reference answer in terms of the vector space models that I used for calculating semantic similarity. This shows that \cite{wolfe1998}'s reference answer approach may not be ideal for grading student answers by means of vector space models.
